---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Kwok, James"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME
str(mydata)
```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)
table(mydata$CLASS, mydata$RINGS)
#boxplot(mydata$LENGTH, outline = TRUE)
#boxplot(mydata$DIAM  , outline = TRUE)
#boxplot(mydata$HEIGHT  , outline = TRUE)
#boxplot(mydata$WHOLE   , outline = TRUE)
#boxplot(mydata$SHUCK    , outline = TRUE)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**





***Answer: (The majority of the variables in the dataset are numerical, with decimal values, with the exception of the "ring" variable, which is an integer. Based on the boxplot, the "whole" and "shuck" variables exhibit potential right skewness.To identify potential outliers, boxplots were created for several variables including "length", "diam", "height", "whole", and "shuck". For "length", any values below 4 are considered outliers. For "diam", any values below 3 are considered outliers. For "height", there are only 2 outliers below 0.5. For "whole", any values above approximately 270 are considered outliers. Finally, for "shuck", any values above 130 are considered outliers.)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
#?addmargins
# Create table of counts
count_table <- table(mydata$SEX, mydata$CLASS)

# Add margins to table
count_table_with_margins <- addmargins(count_table)

# Print table with margins
count_table_with_margins



# Create barplot of data (ignoring marginal totals)
barplot(count_table, beside = TRUE, legend.text = TRUE, args.legend = list(title = "CLASS"), xlab = "SEX" ,ylab="COUNT",col = c( "red", "green","blue"), main= "Abalone Class & Sex")


```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**



***Answer: (In the A1 class of abalone, the sex is still not easy to identify as they are still young and hence are considered as infants. As they continue to age, the sex becomes easier to identify. In the A3 to A5 classes, the count of infants decreased dramatically and the proportion of males and females became evenly divided. The total count also decreased significantly in A4 and A5, which could be due to commercial harvesting.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}

# Set seed for reproducibility
set.seed(123)

# Create simple random sample of 200 observations
work <- mydata[sample(1:nrow(mydata), 200),]

# Construct scatterplot matrix of variables 2-6
plot(work[, 2:6])


```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
library(ggplot2)


ggplot(mydata, aes(x = VOLUME, y = WHOLE, color = CLASS)) + 
  geom_point() + 
  labs(x = "Volume", y = "Whole weight", color = "Class")+
       ggtitle("Whole weight, in grams | Volumn, cm\u00B3") +
  theme(plot.title = element_text(hjust = 0.5))


```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
#?abline
# Calculate maximum ratio of SHUCK to WHOLE
max_ratio <- max(mydata$SHUCK / mydata$WHOLE)
#mydata$CLASS <- factor(mydata$CLASS, levels = c( "A1", "A2", "A3", "A4","A5"))
# Create scatterplot with colored points and straight line
plot(mydata$WHOLE, mydata$SHUCK, col = mydata$CLASS, xlab = "Whole weight", ylab = "Shuck weight",main = "Shucked Weight in grams | Whole Weight in grams ")
abline(a = 0, b = max_ratio, lwd = 2, col = "black", lty=2)
legend("bottomright", legend = unique(mydata$CLASS), col = unique(mydata$CLASS), pch = 1)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (In my observation, the shape of both graphs seems similar. As the volume increases in Graph 2A, the whole weight also tends to increase. Similarly, in Graph 2B, as the whole weight increases, the shucked weight also tends to increase. Thus, if the shucked weight increases, it results in an increase in both the whole weight and volume. On average, A1 has the lightest weight, which is understandable since most of them are infants. On the other hand, A5 tends to be the heaviest, which is also expected as they are fully developed compared to A1.)***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a, fig.height=10, fig.width=10}

# Set up 3 x 3 plot layout
par(mfrow = c(3,3))

# Increase size of plotting area and fonts
par(mar = c(5,5,4,2) + 0.5, cex.lab = 1, cex.axis = 1, cex.main = 1)

# Histograms of RATIO for each sex
hist(mydata[mydata$SEX=="F", "RATIO"], xlab = "",col= "red", main = "Female Ratio")
hist(mydata[mydata$SEX=="I", "RATIO"], xlab = "",col= "green", main = "Infant Ratio")
hist(mydata[mydata$SEX=="M", "RATIO"], xlab = "",col= "blue", main = "Male Ratio")

# Boxplots of RATIO for each sex
boxplot(mydata[mydata$SEX=="F", "RATIO"], xlab = "",ylim = c(0,0.3),col= "red", main = "Female Ratio")
boxplot(mydata[mydata$SEX=="I", "RATIO"], xlab = "",ylim = c(0,0.3),col= "green", main = "Infant Ratio")
boxplot(mydata[mydata$SEX=="M", "RATIO"], xlab = "",ylim = c(0,0.3),col= "blue", main = "Male Ratio")

# Q-Q plots of RATIO for each sex
qqnorm(mydata[mydata$SEX=="F", "RATIO"], col="red", ylim = c(0,0.3),main="Female Ratio")
qqline(mydata[mydata$SEX=="F", "RATIO"])
qqnorm(mydata[mydata$SEX=="I", "RATIO"], col="green", ylim = c(0,0.3),main="Infant Ratio")
qqline(mydata[mydata$SEX=="I", "RATIO"])
qqnorm(mydata[mydata$SEX=="M", "RATIO"], col="blue", ylim = c(0,0.3),main="Male Ratio")
qqline(mydata[mydata$SEX=="M", "RATIO"])

# Reset plot parameters
par(mfrow = c(1,1))




```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Based on my observations of the histogram and Q-Q plot, I believe that the distribution of the data is normal. The histogram shows a bell-shaped curve, which is characteristic of a normal distribution, and the Q-Q plot indicates that the data points closely follow the diagonal line, which is another indication of normality.While I have observed some outliers in the graph, they do not appear to be extreme in nature. Therefore, I believe that my assumption of normality is still reasonable and the presence of these outliers is unlikely to significantly affect my analysis.)***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}


# Identify outliers for each sex
f_outliers <- boxplot.stats(mydata$RATIO[mydata$SEX == "F"],coef=1.5)$out
i_outliers <- boxplot.stats(mydata$RATIO[mydata$SEX == "I"],coef=1.5)$out
m_outliers <- boxplot.stats(mydata$RATIO[mydata$SEX == "M"],coef=1.5)$out

# Subset mydata to include only outliers
f_data <- subset(mydata, SEX=="F" & RATIO %in% f_outliers)
i_data <- subset(mydata, SEX=="I" & RATIO %in% i_outliers)
m_data <- subset(mydata, SEX=="M" & RATIO %in% m_outliers)

# Display outliers using kable function
library(knitr)
kable(rbind(f_data, i_data, m_data))


```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (There appear to be approximately five outliers in each sex category. Specifically, the outliers for the "I" group tend to be in class A1, whereas the outliers for the "F" and "M" groups tend to be in classes A2 and A3. Additionally, the volume tends to be larger for the "F" and "M" groups, while it tends to be smaller for the "I" group.)***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
library(ggplot2)


# Create boxplots for VOLUME and WHOLE by CLASS
p1 <- ggplot(mydata, aes(x = CLASS, y = VOLUME)) +
  geom_boxplot(fill = "white", color = "black") +
  labs(x = "Class", y = "VOLUME") +
  scale_fill_grey() +
  scale_color_grey()

p2 <- ggplot(mydata, aes(x = CLASS, y = WHOLE)) +
  geom_boxplot(fill = "white", color = "black") +
  labs(x = "Class", y = "WHOLE") +
  scale_fill_grey() +
  scale_color_grey()

# Create side-by-side scatterplots for VOLUME and WHOLE versus RINGS
p3 <- ggplot(mydata, aes(x = RINGS, y = VOLUME, color = CLASS)) +
  geom_point(color = "black") +
  labs(x = "RINGS", y = "VOLUME") +
  scale_color_manual(values = c("black", "black", "black"))

p4 <- ggplot(mydata, aes(x = RINGS, y = WHOLE, color = CLASS)) +
  geom_point(color = "black") +
  labs(x = "RINGS", y = "WHOLE") +
  scale_color_manual(values = c("black", "black", "black"))

# Combine plots into a single figure

grid.arrange(p1, p2, p3, p4, ncol = 2)




```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (Based on the observed trends in the graphs, it is likely that these variables could be good predictors of age. As class increases, both volume and whole tend to be more towards the mean, which is also observed in the trend of increasing rings with increasing volume and whole values. In other words, all four graphs show a positive trend, indicating a general increase in these variables with increasing age. Therefore, it is reasonable to assume that these variables could be useful in predicting age.)***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
# Compute mean values using aggregate function

mean_Vol <-aggregate(VOLUME ~ SEX + CLASS, data = mydata, mean)
mean_Shuck <-aggregate(SHUCK ~ SEX + CLASS, data = mydata, mean)
mean_Ratio<-aggregate(RATIO ~ SEX + CLASS, data = mydata, mean)



# Create matrices of the mean values
vol_mat <- matrix(mean_Vol$VOLUME, nrow = 3, byrow = FALSE)
shuck_mat <- matrix(mean_Shuck$SHUCK, nrow = 3, byrow = FALSE)
ratio_mat <- matrix(mean_Ratio$RATIO, nrow = 3, byrow = FALSE)



# Label the rows by SEX and columns by CLASS
rownames(vol_mat) <- rownames(shuck_mat) <- rownames(ratio_mat) <- c("Female", "Infant", "Male")
colnames(vol_mat) <- colnames(shuck_mat) <- colnames(ratio_mat) <- c("A1", "A2", "A3", "A4", "A5")


# Print the matrices using kable function

kable(vol_mat, caption = "$volume", position = "topleft", digits = 2) 
kable(shuck_mat, caption = "$shuck", position = "topleft", digits = 2) 
kable(ratio_mat, caption = "$Ratio", position = "topleft", digits = 4) 



```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.height=15, fig.width=4}
par(mar = c(5,5,4,1) , cex.lab = 0.5, cex.axis = 0.5, cex.main = 0.5 )
# Compute the mean values
mean_Ratio <- aggregate(RATIO ~ SEX + CLASS, data = mydata, mean)
mean_Vol <- aggregate(VOLUME ~ SEX + CLASS, data = mydata, mean)
mean_Shuck <- aggregate(SHUCK ~ SEX + CLASS, data = mydata, mean)

# Create the plots
# Create the plots
# Set up the plot parameters
par(mfrow = c(3, 1))
colors <- c("red", "green", "blue") # Define line colors



# Mean VOLUME vs CLASS
interaction.plot(mean_Vol$CLASS, mean_Vol$SEX, mean_Vol$VOLUME, 
                 xlab = "CLASS", ylab = "VOLUME", 
                 main = "Mean VOLUME per CLASS",
                 col = colors, lty = 1,
                 lwd= 2,
                 trace.label="sex",
                 xpd=TRUE) # Set line color and style

# Mean SHUCK vs CLASS
interaction.plot(mean_Shuck$CLASS, mean_Shuck$SEX, mean_Shuck$SHUCK, 
                 xlab = "CLASS", ylab = "SHUCK", 
                 main = "Mean SHUCK per CLASS",
                 col = colors, lty = 1,
                 lwd= 2,
                 trace.label="sex",
                 xpd=TRUE) # Set line color and style

# Mean RATIO vs CLASS
interaction.plot(mean_Ratio$CLASS, mean_Ratio$SEX, mean_Ratio$RATIO, 
                 xlab = "CLASS", ylab = "RATIO", 
                 main = "Mean RATIO per CLASS",
                 col = colors, lty = 1,
                 lwd= 2,
                 trace.label="sex",
                 xpd=TRUE) # Set line color and style

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c, fig.height=5, fig.width=5}
# Subset the data to include only abalones with RINGS < 16
subset_data <- subset(mydata, RINGS < 16)

# Set up the plotting window
par(mfrow = c(2, 2))

# Boxplot of VOLUME by RINGS for infants
boxplot(VOLUME ~ RINGS, data = subset_data[subset_data$SEX == "I", ], ylim = c(0, 1100),
        xlab = "RINGS", ylab = "VOLUME", main = "Infant Volume | Rings" , col= "blue")

# Boxplot of VOLUME by RINGS for adults
boxplot(VOLUME ~ RINGS, data = subset_data[subset_data$SEX != "I", ], ylim = c(0, 1100),
        xlab = "RINGS", ylab = "VOLUME", main = "Adult Volume | Rings", col= "red")

# Boxplot of WHOLE by RINGS for infants
boxplot(WHOLE ~ RINGS, data = subset_data[subset_data$SEX == "I", ], ylim = c(0, 400),
        xlab = "RINGS", ylab = "WHOLE", main = "Infant Whole Weight | Rings", col= "blue")

# Boxplot of WHOLE by RINGS for adults
boxplot(WHOLE ~ RINGS, data = subset_data[subset_data$SEX != "I", ], ylim = c(0, 400),
        xlab = "RINGS", ylab = "WHOLE", main = "Adult Whole Weight | Rings", col= "red")


```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (In terms of growth, infants have certain limits. For example, their whole value unlikely to exceed 200 and volume unlikely to exceed 400. In contrast, for adults, both values could fluctuate significantly. Infants have a smaller range of whole value and volume compared to adults. As adults age and their rings increase, the range of whole value and volume also increases. )***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**The abalone dataset contains mostly numerical variables with decimal values, except for the "ring" variable which is an integer. Boxplots were used to identify potential outliers, and outliers were found for several variables. The sex of abalones in class A1 is difficult to identify as they are considered infants, but it becomes easier as they age. The proportion of males and females becomes more evenly divided in classes A3 to A5, and the total count decreases in A4 and A5, which could be due to commercial harvesting. The shape of the graphs for volume and weight variables in both sexes exhibit similar patterns. Based on the observed trends in the graphs, it is likely that volume and whole weight variables could be good predictors of age. The distribution of the data appears to be normal, and the presence of outliers is not extreme. Overall, it is reasonable to assume that these variables could be useful in predicting age.Lastly,ehe number of rings alone cannot determine the sex of an abalone, but it can help to differentiate between adults and infants based on their whole weight and volume.**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (The age of abalone cannot be accurately determined based on physical measurements alone, as evidenced by the difficulty in age prediction from size alone, as seen in the last graph. The mean size of both infant and adult abalone is similar, making it challenging to determine age based on size alone, especially when excluding outliers. While physical measurements can be useful for age prediction in some contexts, accuracy depends on various factors. The original study found physical measurements to be unreliable predictors of age in abalone, likely due to environmental and other factors that affect growth and development. For example, factors such as temperature, water quality, and food availability can impact abalone growth, and may have a greater influence on age prediction than physical measurements. Additionally, factors like disease, injury, and competition for resources can also affect abalone growth, but are more challenging to measure accurately. Thus, physical measurements alone may not be sufficient for accurate age prediction in abalone. )***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (Before accepting an overall histogram and summary statistics as representative of a population or phenomenon, it is important to ask questions to ensure the data is accurate and unbiased. One should ask about the sample size, sampling method, and potential sources of bias. Additionally, one should look for outliers, unusual patterns, or missing data that could affect the validity of the results. Overall, it is important to approach data with a critical mindset and verify its quality before drawing conclusions.)***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: (Observational studies can present difficulties in analyzing data as they lack the control of experimental studies, making it harder to determine causality. However, they can still provide valuable insights into the relationships between variables and potential risk factors. Observational studies may be used to generate hypotheses for further research and may also be more applicable to real-world situations.)***
